---
title: "Price Analysis for Single Family Housing in Buffalo" 
author: Erik Woyciesjes
subtitle: "2017 - 2020"
---

# Introduction

The purpose of this project is to look at single family housing in the City of Buffalo. Using the assessed values of single family properties (City Assessment Rolls), this project aims to examine the spatial distribution and percent change of single family housing prices across the city of Buffalo. To extend this study, median household prices within neighborhoods will also be calculated to potentially illuminate trends in the changes in price of single family homes across neighborhoods. 

# Materials and methods

[~ 200 words]

Sources: Clear narrative description of the data sources and methods. Includes data from at least two sources that were integrated / merged in R.

Code: The code associated with the project is well organized and easy to follow. Demonstrates mastery of R graphics and functions.

Data: The underlying data are publicly accessible via the web and downloaded/accessed within the Rmd script. If you want to use your own data, you must make it available on a website (e.g. Figshare) so that others are able to re-run your code.

# Packages

```{r, message=F, warning=F}
library(tidyverse)
library(ggplot2)
library(ggmap)
library(maptools)
library(ggthemes)
library(rgeos)
library(broom)
library(plyr)
library(dplyr)
library(grid)
library(gridExtra)
library(reshape2)
library(scales)
library(sp)
library(sf)
library(rgdal)
library(RColorBrewer)
library(kableExtra)
library(leaflet)
knitr::opts_chunk$set(cache=TRUE) 
```

## Data Download & Cleaning
```{r, message=F, warning=F, results='hide'}
# 2017 - 2019 Buffalo Assessment Roll
Parcel17 <- read.csv(file = "https://raw.githubusercontent.com/geo511-2020/geo511-2020-project-erikwoyc/master/2017-2018_Assessment_Roll.csv")
SingleFam_propclass <- c("210", "215", "240", "241", "250", "270")
Buffalo_17 <- filter(Parcel17, PROPERTY.CLASS %in% SingleFam_propclass)

# 2019 - 2020 Buffalo Assessment Roll
Parcel20 <- read.csv(file = "https://raw.githubusercontent.com/geo511-2020/geo511-2020-project-erikwoyc/master/2019-2020_Assessment_Roll.csv")
SingleFam_propclass <- c("210", "215", "240", "241", "250", "270")
Buffalo_20 <- filter(Parcel20, PROPERTY.CLASS %in% SingleFam_propclass)

# Neighborhood Shapefile
Neighborhood_URL <- "https://data.buffalony.gov/api/geospatial/q9bk-zu3p?method=export&format=GeoJSON"
Buffalo_Neighborhoods <- st_read(dsn = Neighborhood_URL)
Buffalo_sp <- as_Spatial(Buffalo_Neighborhoods)

```
#Snapshot of Assessment Data from 2017 - 2018
```{r, message=F, warning=F}
Parcel17 %>% 
  slice(1:10) %>% 
  kable(digits=2,align="c") %>% 
  kable_styling(bootstrap_options = 
                  c("striped", "hover", "condensed", "responsive")) 
```

# Data Exploration

```{r, message=F, warning=F}
# 2017 - 2018 Single Family Housing Price Histogram
Plot_2017 <- ggplot(data = Buffalo_17, mapping = aes(x = TOTAL.VALUE)) + 
  geom_histogram() + xlab("Total Property Value($)") + ylab("Count") +
  scale_fill_manual(values="lightblue") + theme_few() +
  labs(x="Total Value($)", y="Count", title="Distribution of Buffalo Home Prices",
       subtitle="Single Family Property Prices (2017 - 2018)", 
       caption="Source: Buffalo Open Data") + scale_x_continuous() + scale_y_continuous()
plot(Plot_2017)
```


```{r, message=F, warning=F}
# 2019 - 2020 Single Family Housing Price Histogram
Plot_2019 <- ggplot(data = Buffalo_20, mapping = aes(x = TOTAL.VALUE)) + 
  geom_histogram() + xlab("Total Property Value($)") + ylab("Count") +
  scale_fill_manual(values="lightblue") + theme_few() +
  labs(x="Total Value($)", y="Count", title="Distribution of Buffalo Home Prices",
       subtitle="Single Family Property Prices (2019 - 2020)", 
       caption="Source: Buffalo Open Data") + scale_x_continuous() + scale_y_continuous()
plot(Plot_2019)
```


# Buffalo Base Map
```{r, message=F, warning=F}
#Buffalo Bounding Box
Buffalo_bbox <- Buffalo_sp@bbox

# Download the basemap
basemap <- get_stamenmap(
  bbox = Buffalo_bbox,
  zoom = 13,
  maptype = "toner-lite")

# View Map
BFMap <- ggmap(basemap) + 
  labs(title="Buffalo Basemap")
BFMap
```

# 2017 - 2018 Assessment Roll Plot
```{r, message=F, warning=F}
SingleFam17 <- ggmap(basemap) +
  geom_point(data = Buffalo_17, aes(x = LONGITUDE, y = LATITUDE, color = TOTAL.VALUE), 
             size = .025, alpha = 0.7) +
  scale_color_gradient("Single Family Home Price", low = "light green", high = "dark green", trans="log",
                       labels = scales::dollar_format(prefix = "$")) +
  labs(title="Distribution of Buffalo Home Prices",
       subtitle="Property Prices (2017 - 2018)",
       caption="Open Data Buffalo")
SingleFam17
```

# 2019 - 2020 Assessment Roll Plot
```{r, message=F, warning=F}
SingleFam20 <- ggmap(basemap) + 
  geom_point(data = Buffalo_20, aes(x = LONGITUDE, y = LATITUDE, color = TOTAL.VALUE), 
             size = .025, alpha = 0.7) +
  scale_color_gradient("Single Family Home Price", low = "light green", high = "dark green", trans="log",
                       labels = scales::dollar_format(prefix = "$")) +
  labs(title="Distribution of Buffalo Home Prices",
       subtitle="Property Prices (2019 - 2020)",
       caption="Open Data Buffalo")
SingleFam20
```

# Percent Change in Single Family Home Price from 2017 - 2020
```{r, message=F, warning=F}
Neighborhood_map <- leaflet() %>%
  setMaxBounds(lng1 = -78.91246, lat1 = 42.82603, lng2 = -78.79504, lat2 = 42.96641) %>%
  addProviderTiles("CartoDB") %>%
  addProviderTiles("Stamen.TonerLines",
                   options = providerTileOptions(opacity = 0.35)) %>%
  addPolygons(data = Buffalo_sp, fillColor = "transparent") %>%
  addCircles(data = Buffalo_20, lng = Buffalo_20$LONGITUDE, lat = Buffalo_20$LATITUDE, 
             color = ~pallete(log(Buffalo_20$TOTAL.VALUE)),
             radius = .05, opacity = 0.5) %>%
  addLegend(pal = pallete, values = ~log(TOTAL.VALUE), labFormat = labelFormat(scales::dollar_format(prefix = "$")))
Neighborhood_map
```

# Regression Analysis
```{r, message=F, warning=F}
## Transform Data for Regression
Buffalo_20$log <- log10(Buffalo_20$TOTAL.VALUE)
View(Buffalo_20)

## Multiple Linear Regression
MLR <- lm(log ~ X..OF.BEDS + YEAR.BUILT +
          TOTAL.LIVING.AREA + X..OF.BATHS + NEIGHBORHOOD,
          data = Buffalo_20)
summary(MLR)
```


# Conclusions

[~200 words]

Clear summary adequately describing the results and putting them in context. Discussion of further questions and ways to continue investigation.

# References

All sources are cited in a consistent manner
